Annotation exploration for SCPCP000006, `predicted.score` threshold `r params$predicted.score_thr`

### 2024-11-07 analyses/cell-type-wilms-tumor-06/notebook/04_annotation_Across_Samples_exploration.Rmd
Fix typos and sample_id loop bug, and the file was also styled

- the `counts`of the endothelial marker VWF = "ENSG00000110799"

- the `counts`of the immune marker PTPRC = "ENSG00000081237"

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")

metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE)

# Create a data frames of all annotations
cell_type_df <- metadata$scpca_sample_id |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {

      # Read in the Seurat object
      srat <- readRDS(
        file.path(result_dir, sample_id, paste0("02b-fetal_kidney_label-transfer_", sample_id, ".Rds"))
      )

@@

`do_Feature_mean` shows heatmap of mean expression of a feature grouped by a metadata.

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature ){

  df <- df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5), title = element_text(size = rel(0.75))) +
   guides(fill=guide_colourbar(title=paste0(feature)))

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_boxplot <- function(df, group.by, feature, split.by ){

  df <- df %>%
    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(df, 
    aes(x = group, y = !!sym(feature), fill = group)
  ) + 
  geom_boxplot(size = 0.5, size.outlier = 0.25) +
  facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) + 
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0), 
    legend.position = "none"
  )

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_densityplot <- function(df, group.by, feature){

  df <- df %>%
    mutate(!!sym(group.by), group  = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(df, 
    aes(x = !!sym(feature),  fill = group)
  ) + 
  geom_density() +
  facet_wrap(vars(!!sym(group.by)), scale = "free_y", ncol = 4) + 
  theme_bw()

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

kidney_count_df <- cell_type_df |>
  # make a new variable that tells us if it's kidney or not
  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney"))  |>
  # count how many kidney or not
  dplyr::count(sample_id, organ_type) |>
  # make the data frame wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
  # add a column for total, and convert other counts to percentages
  dplyr::mutate(
    total = kidney + not_kidney, 
    kidney = round(kidney/total,  4)*100, 
    not_kidney = round(not_kidney/total, 4)*100) |>
  # arrange in descending order of percentage of kidney cells
  dplyr::arrange(desc(kidney))

DT::datatable(kidney_count_df, 
              caption = "percentage of cells mapping to the predicted organ kidney",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

The predicted compartment is the result of the label transfer fron the human fetal kidney atlas: 
[Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`

@@

For each compartment, we sumarized the percentage of cells that do match kidney annotation or not. 
Please note that this table is <u>not</u> sample-specific but contains all samples pooled into one.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

compartment_count_df <- cell_type_df |>
  # make a new variable that tells us if it's kidney or not
  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney"))  |>
  # count how many kidney or not
  dplyr::count(compartment, organ_type) |>
  # make the data frame wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
  # add a column for total, and convert other counts to percentages
  dplyr::mutate(
    total = kidney + not_kidney, 
    kidney = round(kidney/total,  4)*100, 
    not_kidney = round(not_kidney/total, 4)*100) |>
  # arrange in descending order of percentage of kidney cells
  dplyr::arrange(desc(kidney))

DT::datatable(compartment_count_df, 
              caption = "percentage of cells mapping to the predicted organ kidney",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

DT::datatable(compartment_df, 
              caption = "counts of cells in each compartment",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_densityplot( df = cell_type_df,
                    feature = "compartment_score",
                    group.by = "compartment"
                    )
p +  geom_vline(xintercept=params$predicted.score_thr)
```

@@

```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "sample_id")

p + geom_hline(yintercept = params$predicted.score_thr)  + coord_cartesian(ylim = c(0.25,1))

```

@@

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "treatment")

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))

```

@@

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "subdiagnosis")

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))

```

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot( df = cell_type_df,
                    feature = "compartment_score",
                  group.by = "compartment",
                    split.by = "disease_timing")

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25,1))

```

@@

DT::datatable(scores_df, 
              caption = "counts of cells that have a good/poor mapping in each compartment",
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',
                               buttons = c( 'csv', 'excel')))
```

@@

We look at the endothelial marker "ENSG00000110799" = "VWF"

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_pass,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for cells passing the `predicted.score` threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_nopass,
                    feature = "ENSG00000110799",
                    group.by = "compartment") +
  ggtitle("VWF expression averaged by compartment for cells that don't pass the `predicted.score` threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000110799",
                    group.by = "compartment",
                    split.by = "sample_id")  +
  ggtitle("boxplot of VWF expression for all cells")

```

@@

We look at the immune marker "ENSG00000081237" = "PTPRC" alias "CD45"

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_pass,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for cells passing the `predicted.score` threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean( df = cell_type_df_nopass,
                    feature = "ENSG00000081237",
                    group.by = "compartment") +
  ggtitle("PTPRC expression averaged by compartment for cells that are not passing the predicted score threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot( df = cell_type_df,
                    feature = "ENSG00000081237",
                    group.by = "compartment",
                    split.by = "sample_id")  +
  ggtitle("boxplot of PTPRC expression for all cells")

```

@@

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
  geom_point() +
  geom_point(data = cell_type_df_nopass, mapping = aes( x = umap.umap_1, y = umap.umap_2),  shape = 4, color = "black", alpha = 0.3) +
  facet_wrap(facets = ~sample_id)
```

@@

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df_pass, aes( x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2)+
  geom_point() +
  facet_wrap(facets = ~sample_id)

```

@@

The next step of the analysis is to test whether `copyKAT` can help us distinguishing cancer cells from non-malignant cell types.
[`copyKAT`](https://github.com/navinlabcode/copykat) (Copynumber Karyotyping of Tumors) is a computational tool using integrative Bayesian approaches to identify genome-wide aneuploidy in single cells to separate tumor cells from normal cells, using high-throughput sc-RNAseq data. 
The underlying logic for calculating DNA copy number events from RNAseq data is that gene expression levels of many adjacent genes can provide depth information to infer genomic copy number in that region.

>>>>

- the `counts` of the endothelial marker `VWF = "ENSG00000110799"`

- the `counts` of the immune marker `PTPRC = "ENSG00000081237"`

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")

metadata <- read.table(sample_metadata_file, sep = "\t", header = TRUE) |>
  dplyr::filter(seq_unit != "spot")

# Create a data frames of all annotations
cell_type_df <- metadata$scpca_sample_id |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {
      # Read in the Seurat object
      srat <- readRDS(
        file.path(result_dir, sample_id, paste0("02b-fetal_kidney_label-transfer_", sample_id, ".Rds"))
      )

@@

`do_Feature_mean` shows a heatmap of mean expression of a feature grouped by a metadata.

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature) {
  df <- df %>%
    group_by(sample_id, !!sym(group.by)) %>%
    summarise(m = mean(!!sym(feature)))

  p <- ggplot(df, aes(x = sample_id, y = !!sym(group.by), fill = m)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(text = element_text(size = 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5), title = element_text(size = rel(0.75))) +
    guides(fill = guide_colourbar(title = paste0(feature)))

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_boxplot <- function(df, group.by, feature, split.by) {
  df <- df %>%
    mutate(!!sym(group.by), group = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(
    df,
    aes(x = group, y = !!sym(feature), fill = group)
  ) +
    geom_boxplot(size = 0.5, size.outlier = 0.25) +
    facet_wrap(vars(!!sym(split.by)), scale = "free_y", ncol = 4) +
    theme_bw() +
    theme(
      axis.text.x = element_text(size = 12, angle = 0, hjust = 0.5, vjust = 0),
      legend.position = "none"
    )

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_densityplot <- function(df, group.by, feature) {
  df <- df %>%
    mutate(!!sym(group.by), group = factor(!!sym(group.by), levels = c("fetal_nephron", "stroma", "immune", "endothelium")))

  p <- ggplot(
    df,
    aes(x = !!sym(feature), fill = group)
  ) +
    geom_density() +
    facet_wrap(vars(!!sym(group.by)), scale = "free_y", ncol = 4) +
    theme_bw()

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
kidney_count_df <- cell_type_df |>
  # make a new variable that tells us if it's kidney or not
  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney")) |>
  # count how many kidney or not
  dplyr::count(sample_id, organ_type) |>
  # make the data frame wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
  # add a column for total, and convert other counts to percentages
  dplyr::mutate(
    total = kidney + not_kidney,
    kidney = round(kidney / total, 4) * 100,
    not_kidney = round(not_kidney / total, 4) * 100
  ) |>
  # arrange in descending order of percentage of kidney cells
  dplyr::arrange(desc(kidney))

DT::datatable(kidney_count_df,
  caption = "percentage of cells mapping to the predicted organ kidney",
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel")
  )
)
```

@@

The predicted compartment is the result of the label transfer from the human fetal kidney atlas: 
[Stewart et al.](https://www.science.org/doi/10.1126/science.aat5031) created a [human fetal kidney atlas](https://www.kidneycellatlas.org/). 
This reference contains only fetal kidney cells and has been precisely annotated by kidney experts. 
The label transfer have been performed in the notebook `02b_fetal_kidney_reference_Stewart_{sample_id}.Rmd`

@@

For each compartment, we summarized the percentage of cells that do match kidney annotation or not. 
Please note that this table is _not_ sample-specific but contains all samples pooled into one.

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
compartment_count_df <- cell_type_df |>
  # make a new variable that tells us if it's kidney or not
  dplyr::mutate(organ_type = ifelse(organ == "Kidney", "kidney", "not_kidney")) |>
  # count how many kidney or not
  dplyr::count(compartment, organ_type) |>
  # make the data frame wide and use a value of 0 if a count is missing
  tidyr::pivot_wider(names_from = organ_type, values_from = n, values_fill = 0) |>
  # add a column for total, and convert other counts to percentages
  dplyr::mutate(
    total = kidney + not_kidney,
    kidney = round(kidney / total, 4) * 100,
    not_kidney = round(not_kidney / total, 4) * 100
  ) |>
  # arrange in descending order of percentage of kidney cells
  dplyr::arrange(desc(kidney))

DT::datatable(compartment_count_df,
  caption = "percentage of cells mapping to the predicted organ kidney",
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel")
  )
)
```

@@

DT::datatable(compartment_df,
  caption = "counts of cells in each compartment",
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel")
  )
)
```

@@

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_densityplot(
  df = cell_type_df,
  feature = "compartment_score",
  group.by = "compartment"
)
p + geom_vline(xintercept = params$predicted.score_thr)
```

@@

```{r fig.height=15, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot(
  df = cell_type_df,
  feature = "compartment_score",
  group.by = "compartment",
  split.by = "sample_id"
)

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
```

@@

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot(
  df = cell_type_df,
  feature = "compartment_score",
  group.by = "compartment",
  split.by = "treatment"
)

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
```

@@

```{r fig.height=3, fig.width=10, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot(
  df = cell_type_df,
  feature = "compartment_score",
  group.by = "compartment",
  split.by = "subdiagnosis"
)

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
```

@@

```{r fig.height=3, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
p <- do_Feature_boxplot(
  df = cell_type_df,
  feature = "compartment_score",
  group.by = "compartment",
  split.by = "disease_timing"
)

p + geom_hline(yintercept = params$predicted.score_thr) + coord_cartesian(ylim = c(0.25, 1))
```

@@

DT::datatable(scores_df,
  caption = "counts of cells that have a good/poor mapping in each compartment",
  extensions = "Buttons",
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel")
  )
)
```

@@

We look at the endothelial marker `"ENSG00000110799" = "VWF"`

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df,
  feature = "ENSG00000110799",
  group.by = "compartment"
) +
  ggtitle("VWF expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df_pass,
  feature = "ENSG00000110799",
  group.by = "compartment"
) +
  ggtitle("VWF expression averaged by compartment for cells passing the `predicted.score` threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df_nopass,
  feature = "ENSG00000110799",
  group.by = "compartment"
) +
  ggtitle("VWF expression averaged by compartment for cells that don't pass the `predicted.score` threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot(
  df = cell_type_df,
  feature = "ENSG00000110799",
  group.by = "compartment",
  split.by = "sample_id"
) +
  ggtitle("boxplot of VWF expression for all cells")
```

@@

We look at the immune marker `"ENSG00000081237" = "PTPRC"` alias `"CD45"`

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df,
  feature = "ENSG00000081237",
  group.by = "compartment"
) +
  ggtitle("PTPRC expression averaged by compartment for all cells")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df_pass,
  feature = "ENSG00000081237",
  group.by = "compartment"
) +
  ggtitle("PTPRC expression averaged by compartment for cells passing the `predicted.score` threshold")
```

```{r fig.height=6, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_mean(
  df = cell_type_df_nopass,
  feature = "ENSG00000081237",
  group.by = "compartment"
) +
  ggtitle("PTPRC expression averaged by compartment for cells that are not passing the predicted score threshold")
```

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
do_Feature_boxplot(
  df = cell_type_df,
  feature = "ENSG00000081237",
  group.by = "compartment",
  split.by = "sample_id"
) +
  ggtitle("boxplot of PTPRC expression for all cells")
```

@@

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2) +
  geom_point() +
  geom_point(data = cell_type_df_nopass, mapping = aes(x = umap.umap_1, y = umap.umap_2), shape = 4, color = "black", alpha = 0.3) +
  facet_wrap(facets = ~sample_id)
```

@@

```{r fig.height=25, fig.width=20, message=FALSE, warning=FALSE, out.width='100%'}
ggplot(cell_type_df_pass, aes(x = umap.umap_1, y = umap.umap_2, color = compartment), shape = 19, size = 2) +
  geom_point() +
  facet_wrap(facets = ~sample_id)
```

@@

The next step of the analysis is to test whether `copyKAT` can help us distinguishing cancer cells from non-malignant cell types.
[`copyKAT`](https://github.com/navinlabcode/copykat) is a computational tool using integrative Bayesian approaches to identify genome-wide aneuploidy in single cells to separate tumor cells from normal cells, using high-throughput scRNAseq data. 
The underlying logic for calculating DNA copy number events from RNAseq data is that gene expression levels of many adjacent genes can provide depth information to infer genomic copy number in that region.
